---
title: "My title"
subtitle: "My subtitle if needed"
author: 
  - Roy Chan
thanks: "Code and data are available at: https://github.com/chan-roy/sta304_paper1"
date: "`r format(Sys.time(), '%d %B %Y')`"
abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
output:
  bookdown::pdf_document2
toc: FALSE
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(ggplot2)
library(dplyr)
library(tidyverse)
```

# Introduction

You can and should cross-reference sections and sub-sections. For instance, Section \@ref(data). R Markdown automatically makes the sections lower case and adds a dash to spaces to generate labels, for instance, Section \@ref(first-discussion-point). 



# Data

As part of the City of Toronto's Vision Zero Road Safety Plan, the Mobile Watch Your Speed Program uses speed display signs which contain a radar device to detect the speeds of oncoming vehicles and an LED display to show the speed of the vehicle to the driver. The aim of the device is to "remind [motorists] to check their speeds and to obey speed limits" [@citeWYSP]. We note that this program is distinct from automated speed enforcement

## TODO ADD CITATIONS
Our data is a summary of speeds recorded by the speed display signs. The data is obtained from (CITATION). The dataset as retrieved from `opendatatoronto` consists of 6517 rows each corresponding to an installation of the display, with columns indicating the location, period of installation, and a summary of speed observations made over the period of time it was installed. All displays are scheduled to only on weekdays between the hours of 7 AM to 9 PM. For each, we have speed distributed by percentiles, as well as the total count of vehicles recorded travelling at a specific speed in 5 km/h increments.

```{r data, fig.cap="Preview of data", echo = FALSE, message = FALSE}
speed_data <- read_csv(here::here("inputs/data/speed_data.csv"))
print(as_tibble(speed_data), n = 3)
```

Given the dimensions of our dataset, we want to split the data by using `dplyr`

```{r histograms, fig.cap="More bills of penguins", echo = FALSE, fig.width=8, fig.height=4}
# drop column schedule as all displays operate on same schedule
locations <- speed_data %>% select(c('_id':removal_date, min_speed))
percentile_speeds <- left_join(locations, select(speed_data, c('_id', starts_with('pct_'))))
speed_counts <- left_join(locations, select(speed_data, c('_id', starts_with('spd_'))))
```

Talk way more about it. 



# Model

\begin{equation}
Pr(\theta | y) = \frac{Pr(y | \theta) Pr(\theta)}{Pr(y)}  (\#eq:bayes)
\end{equation}

Equation \@ref(eq:bayes) seems useful, eh?

Here's a dumb example of how to use some references: In paper we run our analysis in `R` [@citeR]. We also use the `tidyverse` which was written by @thereferencecanbewhatever If we were interested in baseball data then @citeLahman could be useful. 

We can use maths by including latex between dollar signs, for instance $\theta$.


# Results

# Discussion

## First discussion point

If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this. 

## Second discussion point

## Third discussion point

## Weaknesses and next steps

Weaknesses and next steps should also be included.

\newpage

\appendix

# Appendix {-}


# Additional details


\newpage


# References


